\textbf{Efficient Implementation Strategies for Block Ciphers on ARMv8}\\ {\footnotesize Bachelorarbeit}

### 2023-05-11 Paper/chapters/implementation-strategies.tex
bytesliced implementation packing

One caveat of this approach is the fact that NEON registers cannot be shifted
in their entirety due to the fact bits are not able to cross lanes. This leads
to the problem of being able to shift at most two lanes of 64 bits at once. We
thus need to implement the \texttt{shr(V,n)} and \texttt{shl(V,n)} operations
on our own. This can be done by first extracting the 64-bit lanes $a,b$ out of
$V=b||a$, shifting the lanes individually and finally shifting and ORing the
crossing bits back into the other lane.

@@

A byte-sliced implementation strategy for ARMv8 can be derived from already
existing x86-optimized implementations utilizing the AES-NI advanced encryption
standard instruction set \cite{bcfastimplx86:2013}. NEON itself possesses
cryptographic extensions for finite field arithmetic and AES as well as SHA
calculations. These can be used to produce an accelerated Camellia implementation
due to the algebraic similarity of the AES- and Camellia S-boxes.

@@

>>>>

% TODO change shr/shl implementation to only use vector registers
One caveat of this approach is the fact that NEON registers cannot be shifted
in their entirety due to the fact bits are not able to cross lanes. This leads
to the problem of being able to shift at most two lanes of 64 bits at once. We
thus need to implement the \texttt{shr(V,n)} and \texttt{shl(V,n)} operations
on our own. This can be done by first extracting the 64-bit lanes $a,b$ out of
$V=b||a$, shifting the lanes individually and finally shifting and ORing the
crossing bits back into the other lane.

@@

\subsubsection{Packing}

We encrypt 16 plaintext blocks in parallel so we can fill 16 vector registers
with register $i$ containing all $i$-th bytes. This lends itself well to a NEON
implementation since S-boxes as well as the FL layer, Feistel round and
packing/unpacking functions can be implemented efficiently.

@@

A byte-sliced implementation strategy for the S-box on ARMv8 can be derived
from already existing x86-optimized implementations utilizing the AES-NI
advanced encryption standard instruction set \cite{bcfastimplx86:2013}. NEON
itself possesses cryptographic extensions for finite field arithmetic and AES
as well as SHA calculations. These can be used to produce an accelerated
Camellia implementation due to the algebraic similarity of the AES- and
Camellia S-boxes.

The AES S-box works by multiplicatively inverting $x\in\F{8}$ over GF$(2^8)$ and then
applying an affine transformation $A$:

\[
    s(x):\F{8}\rightarrow\F{8}, x\mapsto A(x^{-1}_{\text{GF}(2^8)})
\]

The Camellia S-box $s_0$ is defined as follows:

\[
    s_0(x):\F{8}\rightarrow\F{8}, x\mapsto h(g(f(x\oplus 0xc5)))\oplus 0x6e
\]

with affine transformations $h,f$ and the multiplicative inversion function $g$
in the composite field GF$((2^4)^2)$. Because Galois fields of equal size are
isomorphic, there are affine isomorphisms $\delta,\delta^{-1}$ between
GF($2^8$) and GF($(2^4)^2$) respectively\cite{satohmorioka:2003}:

\begin{alignat*}{2}
    \delta&:\text{GF}(2^8)&&\rightarrow\text{GF}((2^4)^2) \\
    \delta^{-1}&:\text{GF}((2^4)^2)&&\rightarrow\text{GF}(2^8)
\end{alignat*}

NEON provides the \texttt{AESE} instruction for one round of AES encryption
which works on a 128-bit vector register. By applying the inverse of ShiftRow
to x, we can apply the AES S-box to 16 bytes at once.

\[
    \texttt{AESE}(x)=\text{SubBytes}(\text{ShiftRow(x)})\Leftrightarrow \texttt{AESE}(\text{ShiftRow}^{-1}(x))=\text{SubBytes}(x)
\]

We can then reverse the affine transformation $A$ due to bijectivity and
extract the multiplicative inverse of all 16 vector bytes in GF$(2^8)$.

\[
    x^{-1}_{\text{GF}(2^8)}=A^{-1}(A(x^{-1}_{\text{GF}(2^8)}))=A^{-1}(\text{SubBytes}(x))=A^{-1}(\texttt{AESE}(\text{ShiftRow}^{-1}(x)))
\]

Using the affine isomorphism, we transform this inverse into the inverse in
GF$((2^4)^2)$ which is equal to $g(x)$:

\[
    g(x)=x^{-1}_{\text{GF}((2^4)^2)}=\delta(x^{-1}_{\text{GF}(2^8)})=\delta(A^{-1}(\texttt{AESE}(\text{ShiftRow}^{-1}(x))))
\]

We now redefine $h,f$ as $H,F$ such that they include addition of constants
$0xc5$ and $0x6e$. We can now use the inverse in conjunction with $H$ and $F$
and an additional input transformation to calculate the Camellia S-box:

\begin{align*}
    s_0(x)&=h(g(f(x\oplus 0xc5)))\oplus 0x6e \\
          &=H(g(F(x))) \\
          &=H(\delta(A^{-1}(\texttt{AESE}(\text{ShiftRow}^{-1}(\delta^{-1}(F(x)))))))
\end{align*}

Because of different bit endianness of the matrices representing
$\delta,\delta^{-1}$, we define an additional bit-swapping function:

\[
    S=S^{-1}:\begin{pmatrix}x_0\\x_1\\x_2\\x_3\\x_4\\x_5\\x_6\\x_7\end{pmatrix}\mapsto
    \begin{pmatrix}
        0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 \\
        0 & 0 & 0 & 0 & 0 & 0 & 1 & 0 \\
        0 & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\
        0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 \\
        0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 \\
        0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\
        0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
        1 & 0 & 0 & 0 & 0 & 0 & 0 & 0
    \end{pmatrix}
    \begin{pmatrix}x_0\\x_1\\x_2\\x_3\\x_4\\x_5\\x_6\\x_7\end{pmatrix}
\]

The final equation serving as the basis for our implementation then becomes

\[
    s_0(x)=S^{-1}(H(\delta(S(A^{-1}(\texttt{AESE}(\text{ShiftRow}^{-1}(S^{-1}(\delta^{-1}(F(S(x))))))))))
\]

While appearing to be computationally too intensive for any performance gains,
we shall notice that all of the transformations are affine and can therefore be
combined into a single operation by simple matrix multiplication. We combine
them into a pre-filter function $\theta_0$ and a post-filter function $\psi_0$:

\begin{align*}
    \theta_0(x)&=S^{-1}(\delta^{-1}(F(S(x)))) \\
    \psi_0(x)&=S^{-1}(H(\delta(S(A^{-1}(x)))))
\end{align*}

This simplifies our final equation:

\[
    s_0(x)=\psi_0(\texttt{AESE}(\text{ShiftRow}^{-1}(\theta_0(x))))
\]

The other S-boxes $s_1,s_2,s_3$ are defined in terms of $s_0$ and rotations. We
can include these rotations by constructing additional S-box-specific filters
$\theta_3,\psi_1,\psi_2$:

\begin{alignat*}{2}
    s_1(x)&=s_0(x)\lll_1&&=\psi_1(\texttt{AESE}(\text{ShiftRow}^{-1}(\theta_0(x)))) \\
    s_2(x)&=s_0(x)\ggg_1&&=\psi_2(\texttt{AESE}(\text{ShiftRow}^{-1}(\theta_0(x)))) \\
    s_3(x)&=s_0(x\lll_1)&&=\psi_0(\texttt{AESE}(\text{ShiftRow}^{-1}(\theta_3(x))))
\end{alignat*}

\subsubsection{Fast matrix multiplication}

Efficient application of the filter functions is essential to performance and
can be implemented in parallel by use of the \texttt{TBL} instruction. First
note that matrix-vector multiplication $\F{8\times 8}\cdot\F{8}\rightarrow
\F{8}$ can be decomposed into two multiplications $\F{8\times
4}\cdot\F{4}\rightarrow \F{8}$ with subsequent addition of the results:

\[
    \begin{pmatrix}
        1 & 1 & 1 & 0 & 1 & 1 & 0 & 1 \\
        0 & 0 & 1 & 1 & 0 & 0 & 1 & 0 \\
        1 & 1 & 0 & 1 & 0 & 0 & 0 & 0 \\
        1 & 0 & 1 & 1 & 0 & 0 & 1 & 1 \\
        0 & 0 & 0 & 0 & 1 & 1 & 0 & 0 \\
        1 & 0 & 1 & 0 & 0 & 1 & 0 & 0 \\
        0 & 0 & 1 & 0 & 1 & 1 & 0 & 0 \\
        1 & 0 & 0 & 0 & 0 & 1 & 1 & 0 \\
    \end{pmatrix}\cdot
    \begin{pmatrix}x_0\\x_1\\x_2\\x_3\\x_4\\x_5\\x_6\\x_7\end{pmatrix}=
    \begin{pmatrix}
        1 & 1 & 0 & 1 \\
        0 & 0 & 1 & 0 \\
        0 & 0 & 0 & 0 \\
        0 & 0 & 1 & 1 \\
        1 & 1 & 0 & 0 \\
        0 & 1 & 0 & 0 \\
        1 & 1 & 0 & 0 \\
        0 & 1 & 1 & 0 \\
    \end{pmatrix}\cdot
    \begin{pmatrix}x_0\\x_1\\x_2\\x_3\end{pmatrix}\oplus
    \begin{pmatrix}
        1 & 1 & 1 & 0 \\
        0 & 0 & 1 & 1 \\
        1 & 1 & 0 & 1 \\
        1 & 0 & 1 & 1 \\
        0 & 0 & 0 & 0 \\
        1 & 0 & 1 & 0 \\
        0 & 0 & 1 & 0 \\
        1 & 0 & 0 & 0 \\
    \end{pmatrix}\cdot
    \begin{pmatrix}x_4\\x_5\\x_6\\x_7\end{pmatrix}
\]

We can then tabulate the decomposed multiplications since each input now
consists of only four bits which allows us to fit the 16 possible 8-bit results
into a single vector register. A single matrix multiplication is then executed
for 16 bytes in parallel by use of two \texttt{TBL} and one XOR instruction with
an additional AND and SHR operation for masking off the lower/upper 4 bits.
